---
title: "Beyond Correlation: A Practical Introduction to Causal Inference in Observational Social Psychology"
subtitle: "SASP 2025 Pre-conference Workshop"
author:
  - name: "Joseph Bulbulia"
    id: "JB"
    orcid: "0000-0002-5861-2056"
    email: "joseph.bulbulia@vuw.ac.nz"
    corresponding: true
    affiliations:
      - name: "Victoria University of Wellington, New Zealand"
        id: "VUW"
format:
  html:
    highlight-style: dracula
    code-fold: true
    code-overflow: scroll
    code-line-numbers: true
execute:
  echo: false
  warning: false
  error: false
editor_options:
  chunk_output_type: console
---

::: callout-note
### Workshop location
**Date:** Wednesday, 19 November 2025, 09:00–16:00  

**Location:** University of Melbourne (Parkville campus), Redmond Barry Building (Building 115), Level 5 Room 516  

**Format:** Registration via SASP abstract submission; in-person. Before lunch we'll cover average treatment effects.  This module is self-sufficient, and will culminate with student presentations. If you just want to get a sense of causal inference, and learn about causal graphs, the morning will be sufficient. After lunch we'll get into estimation of heterogeneous treatment effects, and walk through a data exercise that will show you how to estimate such effects, as well as average treatment effects. You will estimate effects using **causal machine learning**.  Building knowledge and skills in this area will be empowering.

**Preparation**: Bring your curiosity and attention. If you want to do the data exercise, download R and Rstudio (or your favourite IDE) and then download the workshop R package. (Make sure you install the package first, and run through the scripts as we will not pause to work on your package installation). Bring a writing instrument and some paper, or software that allows free hand writing, as we'll draw graphs together. 
:::


::: {.callout-important}
## Links to Lectures 
In the weeks before the workshop, I'll be **simplifying the background lectures** and adding some background videos.  Updates will be described here. 
:::

## Venue Location

We'll meet on Level 5 of the Redmond Barry Building (Building 115). See the campus map for map for detail by clicking the arrow in the centre of the OpenStreetMap. 

```{r}
#| echo: false
#| message: false
#| warning: false
#| column: body-outset

library(leaflet)

leaflet() %>%
  addTiles() %>%
  addMarkers(
    lng = 144.9617,
    lat = -37.7971,
    popup = "Redmond Barry Building (Building 115)<br>L5 Room 516<br><a href='https://maps.unimelb.edu.au/parkville/building/115' target='_blank'>View on campus map</a>"
  )
```

## Why is causal inference important? 

Suppose you observe that psychology graduate students who own expensive espresso machines publish more papers. Should your department invest in high-end coffee equipment for all incoming students?

Probably not. This is the difference between prognostic and causal knowledge.

Observational psychology typically recovers prognostic information -- which factors predict which others. The word 'predictor' is engrained in our vocabulary. But predictions don't tell us what would happen if we intervened. Owning an expensive coffee machine predicts publication success, but giving someone an expensive machine wouldn't cause their productivity to increase. The machine is likely a marker for other factors: being further along in candidature, having more disposable income, or possessing pre-existing caffeinated ambition. Of course the machines might help, but how much?  


This distinction between **prediction** and **causation** has serious implications for intervention design:

1) We might observe that adolescents with higher self-esteem have better academic outcomes. But if we were to intervene to boost self-esteem directly (through positive affirmations or praise), would grades improve? Or is self-esteem mostly a marker for other causal factors like stable family environments or prior academic success?  

2) We might find that people who practice mindfulness meditation report lower anxiety. But does meditation cause reduced anxiety, or are less anxious people simply more likely to maintain a meditation practice -- i.e. to sit, for long periods of time, still?

3) We might observe that children who attend preschool show better social skills. But before recommending universal preschool, we need to ask: would sending any child to preschool produce these benefits, or do families who can afford and prioritise preschool differ in ways that independently foster social development?  

**Prognostic models tell us what to expect. Causal models tell us what we can change. For psychologists aiming to improve human welfare, that distinction is everything.**

## The workshop

This full-day, practice-oriented workshop introduces modern causal inference for observational data, with an emphasis on discovering for whom effects are strongest (afternoon).

We will:

- Unveil a **causal workflow** that starts with asking a clearly formulated causal question. 
- Before answering a causal question we must consider and explicate **identification assumptions**. We will consider these assumptions and introduce graphical tools for addressing them. You will discover that clarifying assumptions is not a statistical task. 
- Only after stating our questions and assumptions can we turn to the data and its analysis. By the end of the workshop, you will conduct an analysis using doubly robust machine learning -- which has many advantages over standard estimators.  
- After estimation there is communication.  I'll describe how to graphical tools to clarify results audiences with applied interests.  


### Objectives

You will leave with:

- An understanding of how to ask, and answer, causal questions.
- Annotated R code for simulating and estimating average and heterogeneous treatment effects.  
- A curated reading list for self-study.  
- Examples of how to communicate causal results to academic, policy, and organisational audiences.

## My assumptions about you.

The workshop assumes familiarity with regression and R, but no prior training in causal inference.

::: callout-tip
### A note on pace
Please do not worry if the math or code looks intimidating at first glance. We'll slow everything down, build each idea together from first principles, and keep the focus on intuition before symbols. Bring your curiosity. If you want to do the data exercise, bring a computer pre-loaded with the workshop R (see👇).  By the end of the day you'll understand that causal workflow is manageable, and even fun, when we unpack it step by step.
:::

## Setup

1. Install the current version of R (free) from <https://cran.r-project.org/>

2. Install RStudio Desktop (free) from <https://posit.co/download/rstudio-desktop/>  

(Or pick your favourite IDE -- I use neovim Lazy -- the code doesn't care.)

3. Install the workshop packages:

```r
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

devtools::install_github("go-bayes/causalworkshop")

# optional packages for extended demonstrations
devtools::install_github("go-bayes/margot")
```

Once installed, run:

```r
library(causalworkshop)
check_workshop_prerequisites()
```

Download the R scripts:

```r
library(causalworkshop)

get_workshop_scripts()  # copies scripts into ./workshop-scripts
list_workshop_scripts()
```

Work through scripts 01–05 **Prior to the workshop** (06 is optional).

## Workshop materials

- The lectures are somewhat technical at the moment. I'll be simplifying them before the workshop and posting YouTube videos prior to the workshop. Stay tuned here for updates.
- Student presentations will be posted here **after** the workshop.  
- Previous talks and examples remain available for reference; we will link to them from the relevant sections.


## Agenda

```{r}
#| label: tbl-schedule
#| tbl-cap: "Schedule and Learning Goals"
#| eval: true

library(tidyverse)
library(kableExtra)

schedule <- tibble(
  Time = c(
    "09:00–09:45",
    "09:45–10:30",
    "10:30–11:00",
    "11:00–11:30",
    "11:30–12:00",
    "12:00–12:30",
    "12:30–13:45",
    "13:45–14:30",
    "14:30–15:00",
    "15:00–16:00"
  ),
  Session = c(
    "How to ask a causal question",
    "The causal workflow: from question to answer",
    "Causal diagrams and the identification problem",
    "Coffee break & discussion (you might need to bring a your own coffee)",
    "Worked example: employer gratitude and worker wellbeing (Zahle Wisely)",
    "Worked example: religious service attendance and personality change (Hannah Robinson)",
    "Lunch",
    "Beyond averages: heterogeneous treatment effects and policy trees",
    "Revisiting our examples: policy tree analyses",
    "Hands-on demonstration: formulating questions and estimating effects in R"
  ),
  `Learning goals & activities` = c(
    "State precise causal questions with the potential outcomes framework; focus on the average treatment effect (ATE).",
    "Trace the causal workflow: define populations, build causal diagrams, assess assumptions, analyse data, conduct sensitivity analysis, and communicate results.",
    "Use causal diagrams (DAGs) to identify bias from confounding, colliders, and mediators.",
    "Informal space for questions, clarifying assumptions, and networking.",
    "Presenter: Zahle Wisely. Apply the causal workflow to estimate the effects of employer gratitude on employee wellbeing; Q&A follows.",
    "Presenter: Hannah Robinson. Apply the causal workflow to estimate the effects of religious service attendance on personality; Q&A follows.",
    "Break.",
    "Introduce heterogeneous treatment effects (HTE), conditional ATEs, and policy trees; discuss when effects differ across people.",
    "Analyse heterogeneity in Zahle and Hannah's case studies; interpret tree-based decision rules useful for obtaining *policies*.",
    "R demonstration using causal forests; interpret simulated ATEs and HTEs, graphs, reporting & etc."
  )
)

schedule %>%
  kbl(col.names = c("Time", "Session", "Learning goals & activities")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = TRUE
  )
```

## Contact

Questions before the workshop? Email me at [joseph.bulbulia@vuw.ac.nz](mailto:joseph.bulbulia@vuw.ac.nz). 
