---
title: "Reporting Causal Inference Results"
subtitle: "How to Present Your Analysis Findings"
date: "May 2025"
format:
  html:
    code-overflow: scroll
    code-tools:
      source: true
      toggle: false
    toc: true
    toc-depth: 3
    number-sections: true
warning: false
error: false
---

## Introduction

This guide demonstrates how to report results from causal inference analyses following the 9-step causal inference workflow. We use the workshop example of religious service → cooperation to illustrate best practices for communicating causal findings.

::: {.callout-important}
## Focus
This tutorial covers **result presentation** following rigorous causal inference methodology. Learn how to interpret and communicate your causal findings using both standardised and data-scale effects.
:::

## The 9-Step Causal Inference Workflow

Before reporting results, ensure you've addressed these foundational steps:

### Steps 1-3: Problem Definition
1. **Well-defined treatment**: "Weekly religious service attendance (≥1 time per week vs. <1 time per week)"
2. **Well-defined outcome**: "Cooperative behaviour measured using standardised scales one year after intervention"  
3. **Target population**: "New Zealand adults aged 18+ as represented in NZAVS Time 10 weighted by census demographics"

### Steps 4-6: Identification Strategy
4. **Exchangeability**: Conditional independence achieved through rich baseline covariate control
5. **Consistency**: Binary exposure ensures well-defined intervention versions
6. **Positivity**: Verified through transition tables showing exposure variation across covariates

### Steps 7-9: Implementation
7. **Measurement validity**: Validated scales with known psychometric properties
8. **Attrition handling**: Inverse probability of censoring weights (IPCW) for panel attrition
9. **Transparent reporting**: Complete documentation of decisions and assumptions

::: {.callout-tip}
## Target Trial Emulation
Frame your causal question as: "How would outcomes change if we intervened to set everyone's religious service attendance to weekly rather than less-than-weekly, conditional on baseline characteristics?"
:::

## Reporting Average Treatment Effects

### Standard ATE Table

When reporting average treatment effects, include these key elements:

**Table 1: Average Treatment Effects of Weekly Religious Service on Cooperation**

| Outcome | Estimate | 95% CI | E-value | Interpretation |
|---------|----------|---------|---------|----------------|
| Charitable Donations (log) | 0.12 | [0.08, 0.16] | 1.8 | Small positive effect |
| Volunteering Hours (log) | 0.09 | [0.05, 0.13] | 1.6 | Small positive effect |
| Social Support | 0.15 | [0.11, 0.19] | 2.1 | Moderate positive effect |
| Sense of Belonging | 0.18 | [0.14, 0.22] | 2.4 | Moderate positive effect |

### Key Reporting Elements

1. **Effect sizes**: Report in standard deviation units for continuous outcomes
2. **Confidence intervals**: Show uncertainty around estimates
3. **E-values**: Indicate robustness to unmeasured confounding
4. **Sample size**: Total analysed after exclusions/weighting

### Example Results Text

> "Weekly religious service attendance showed positive causal effects across all cooperation measures. The largest effects were observed for social outcomes: sense of belonging (β = 0.18, 95% CI: 0.14-0.22) and social support (β = 0.15, 95% CI: 0.11-0.19). Behavioural cooperation showed smaller but significant effects: charitable donations (β = 0.12, 95% CI: 0.08-0.16) and volunteering (β = 0.09, 95% CI: 0.05-0.13). All effects were robust to moderate unmeasured confounding (E-values > 1.6)."

## Reporting Heterogeneous Treatment Effects

### RATE Results Summary

When you find heterogeneity, report it systematically:

**Table 2: Heterogeneity Assessment Using RATE Metrics**

| Outcome | RATE-AUTOC | p-value | RATE-Qini | p-value | Evidence |
|---------|------------|---------|-----------|---------|----------|
| Social Support | 0.12 | 0.003 | 0.08 | 0.012 | Strong |
| Belonging | 0.15 | 0.001 | 0.11 | 0.004 | Strong |
| Charitable Donations | 0.06 | 0.089 | 0.04 | 0.156 | Moderate |
| Volunteering | 0.03 | 0.234 | 0.02 | 0.445 | Weak |

### Example Heterogeneity Text

> "We found substantial heterogeneity in treatment effects for social outcomes (RATE-Qini > 0.08, p < 0.05), but limited heterogeneity for behavioural outcomes. This suggests that while religious service benefits most people socially, individual responses vary considerably in magnitude."

## Reporting Policy Tree Results

### Subgroup Identification with Data-Scale Effects

Present policy trees with both standardised and original scale interpretations:

```{r}
#| eval: false
# Example: Convert standardised effects to original scale for donations
log_mean_donations <- mean(original_df$log_charity_donate, na.rm = TRUE)
log_sd_donations <- sd(original_df$log_charity_donate, na.rm = TRUE)

# Create mapping for common effect sizes
donation_mapping <- margot_back_transform_log_z(
  log_mean = log_mean_donations,
  log_sd = log_sd_donations,
  z_scores = c(0.08, 0.22, 0.28),  # effect sizes from subgroups
  label = "annual_donations_nzd"
)
print(donation_mapping)
```

**High-Response Subgroups for Charitable Donations:**

1. **Older adults with high agreeableness** (Age > 45 & Agreeableness > +1 SD)
   - **Standardised effect**: β = 0.28 (95% CI: 0.21-0.35)
   - **Data-scale effect**: NZ$847 additional annual donations (95% CI: NZ$635-1,058)
   - **Sample proportion**: 23%

2. **Parents with medium conscientiousness** (Parent = Yes & Conscientiousness > 0)
   - **Standardised effect**: β = 0.22 (95% CI: 0.16-0.28)
   - **Data-scale effect**: NZ$665 additional annual donations (95% CI: NZ$484-846)
   - **Sample proportion**: 31%

3. **All others**
   - **Standardised effect**: β = 0.08 (95% CI: 0.04-0.12)
   - **Data-scale effect**: NZ$242 additional annual donations (95% CI: NZ$121-363)
   - **Sample proportion**: 46%

### Example Policy Tree Text with Data-Scale Interpretation

> "Policy tree analysis identified two subgroups with enhanced treatment response for charitable donations. Older adults (45+) with high agreeableness showed the largest increase (β = 0.28 SD, equivalent to NZ$847 additional annual donations), representing 23% of the sample. Parents with medium-to-high conscientiousness also showed strong responses (β = 0.22 SD, equivalent to NZ$665 additional annually), comprising 31% of the sample. In practical terms, targeted interventions toward these subgroups could generate 2.8-3.5 times more charitable giving than population-wide approaches."

## Sensitivity Analysis Reporting

### E-value Interpretation

Always interpret E-values in context:

> "To assess robustness to unmeasured confounding, we calculated E-values for all estimates. The observed effect on sense of belonging (E-value = 2.4) would require an unmeasured confounder associated with both religious service and belonging by a risk ratio of 2.4 each to explain away the result. This represents a strong confounder—stronger than most demographic or personality variables in our model."

### What E-values Mean

- **E-value ≥ 2.0**: Robust to strong confounding
- **E-value 1.5-2.0**: Robust to moderate confounding  
- **E-value < 1.5**: Vulnerable to weak confounding

## Methods Section Template Following 9-Step Workflow

### Target Trial Emulation

```{r}
#| eval: false
# Complete methods section template:

"We emulated a target trial to estimate the causal effect of religious 
service attendance on cooperative behaviour.

**Treatment Definition**: We defined the treatment as weekly religious 
service attendance (≥1 time per week) compared to less frequent attendance 
(<1 time per week), assessed during Time 11 (2019) of the New Zealand 
Attitudes and Values Study.

**Outcome Definition**: Cooperative behaviour was measured using four 
validated scales one year later (Time 12, 2020): charitable donations 
(log-transformed annual giving), volunteering hours (log-transformed 
weekly hours), social support (7-point scale), and sense of belonging 
(7-point scale). All outcomes were z-standardised for analysis.

**Target Population**: New Zealand adults aged 18+ as represented in NZAVS 
Time 10, weighted by census demographics for age, gender, and ethnicity 
to enhance population representativeness.

**Eligibility Criteria**: Participants were included if they provided 
baseline religious service data (Time 10) and were not lost to follow-up 
due to COVID-related measurement error. Of 20,000 eligible participants, 
15,742 remained after applying exclusion criteria and censoring weights.

**Causal Identification**: We assumed conditional exchangeability given 
32 baseline covariates including demographics, personality traits, health 
indicators, and prior cooperation levels. Positivity was verified through 
transition tables showing exposure variation across covariate strata.

**Statistical Analysis**: We estimated causal effects using generalized 
random forests with sample splitting to prevent overfitting. The analysis 
employed 1,000 trees with minimum node size 20. We applied two-stage 
inverse probability of censoring weights to address panel attrition: 
stage 1 for baseline-to-exposure dropout, stage 2 for exposure-to-outcome 
dropout, combined with original sampling weights.

**Heterogeneity Assessment**: Treatment effect heterogeneity was evaluated 
using Rank Average Treatment Effect (RATE) metrics with false discovery 
rate correction. For outcomes showing significant heterogeneity (p < 0.05), 
we fitted policy trees with maximum depth 2 to identify high-response 
subgroups.

**Sensitivity Analysis**: Robustness to unmeasured confounding was assessed 
using E-values, quantifying the minimum strength of association an unmeasured 
confounder would require with both exposure and outcome to explain away 
observed effects."
```

## Key Reporting Principles

### Do Report

✅ **Both standardised and data-scale effects**  
✅ **Effect sizes with confidence intervals**  
✅ **Sample sizes after exclusions and weighting**  
✅ **E-values for sensitivity analysis**  
✅ **Clear practical interpretation of effect magnitudes**  
✅ **Subgroup sizes and effect magnifications**  
✅ **Target trial framework and causal question**  
✅ **Explicit treatment and outcome definitions**  

### Don't Report

❌ **Model coefficients without interpretation**  
❌ **P-values alone without effect sizes**  
❌ **Only standardised effects for policy-relevant outcomes**  
❌ **Technical details that obscure main findings**  
❌ **Causal claims beyond your identification strategy**  
❌ **Results without addressing the 9-step workflow**

## Figure Presentation

### ATE Plot Standards

- Use forest plots with confidence intervals
- Order by effect magnitude or E-value
- Include sample sizes
- Use consistent colour coding

### Policy Tree Plots

- Show decision rules clearly
- Include sample proportions in each node
- Use readable font sizes
- Provide plain-language interpretation

## Conclusion

Effective reporting of causal inference results requires:

1. **Clear effect size interpretation** in practical terms
2. **Honest uncertainty quantification** with confidence intervals  
3. **Sensitivity analysis** using E-values
4. **Meaningful subgroup analysis** when heterogeneity exists
5. **Accessible language** that explains technical concepts

The goal is communicating actionable insights, not demonstrating methodological sophistication.

::: {.callout-tip}
## Remember
Your analysis is only as good as how clearly you can communicate the findings. Focus on what decision-makers need to know, not just what the statistical tests show.
:::

## Example Complete Results Section with Data-Scale Effects

> **Results**
> 
> **Average Treatment Effects**: Among 15,742 New Zealand adults followed over three years, weekly religious service attendance caused small to moderate increases in cooperative behaviour. Social cooperation showed the strongest effects: sense of belonging increased by 0.18 standard deviations (95% CI: 0.14-0.22), equivalent to a 1.3-point increase on the 7-point scale, and social support increased by 0.15 SD (95% CI: 0.11-0.19), equivalent to a 1.1-point increase. Behavioural cooperation showed smaller but meaningful effects: charitable donations increased by 0.12 SD (95% CI: 0.08-0.16), equivalent to NZ$364 additional annual giving, and volunteering increased by 0.09 SD (95% CI: 0.05-0.13), equivalent to 1.7 additional hours weekly.
> 
> **Heterogeneous Treatment Effects**: Treatment effects varied substantially across individuals for social outcomes (RATE-Qini > 0.08, p < 0.05) but showed little heterogeneity for behavioural outcomes. Policy tree analysis identified high-response subgroups for charitable donations: older adults (45+) with high agreeableness showed the largest response (β = 0.28 SD, equivalent to NZ$847 additional annually), while parents with medium conscientiousness showed strong responses (β = 0.22 SD, equivalent to NZ$665 additional annually). Together, these subgroups comprised 54% of the sample and showed 2.8-3.5 times larger effects than the general population.
> 
> **Sensitivity Analysis**: All findings were robust to unmeasured confounding, with E-values ranging from 1.6 to 2.4. Even the smallest effect (volunteering, E-value = 1.6) would require an unmeasured confounder associated with both religious attendance and volunteering by a risk ratio of 1.6 each to explain away the result. The largest effect (belonging, E-value = 2.4) would require substantially stronger confounding to nullify.

::: {.callout-tip}
## Best Practice Summary

**Complete causal inference reporting requires:**

1. **9-Step Workflow Compliance**: Address all foundational steps
2. **Target Trial Framework**: Frame as hypothetical intervention  
3. **Dual-Scale Reporting**: Both standardised and data-scale effects
4. **Heterogeneity Assessment**: RATE metrics and policy trees when appropriate
5. **Sensitivity Analysis**: E-values for robustness evaluation
6. **Clear Practical Interpretation**: Policy-relevant effect magnifications

Remember: The goal is actionable insights for decision-makers, not just statistical significance.
:::